|

### 2024-10-14 posts/BarthPapageorgeThom/BPT.qmd
Added sky_add_FilePointer.html to apply, if want to, sky_add.css to all quarto posts
Added sky_add_FilePointer.html to apply, if want to, seiro.css to all quarto posts

title: "Genetic endowments and wealth inequality"

@@

  Journal of Political Economy, 2020, vol. 128, No.4

@@

    theme: [sky, sky-add.scss]

@@

    include-in-header: ../sky_add_FilePointer.html
    include-after-body: ../toc-add.html

@@

  - imagify
imagify:
  header-includes: |
    \usetikzlibrary{calc}
    \usetikzlibrary{positioning}
    \usetikzlibrary{math}
    \usetikzlibrary{decorations.pathmorphing}
    \usetikzlibrary{shapes}
    \usetikzlibrary{arrows}
    \usetikzlibrary{backgrounds}
    \usetikzlibrary{fit}
    \usepackage{pgfplots}
  pdf-engine: lualatex

@@

@@

Genome

@@

::::{.fragment}
![Chromosone...DNA](DNAMacrostructure.jpg)

::: {style="font-size: 70%;"}
<https://courses.lumenlearning.com/suny-ulster-ap1/chapter/the-nucleus-and-dna-replication/>
:::
::::

@@

::: {style="font-size: 70%;"}
<https://www.haibunda.com/bundapedia/20230214003129-210-297249/deoxyribonucleic-acid--dna->
:::
::::

@@

<https://mammothmemory.net/biology/dna-genetics-and-inheritance/dna-base-pairing/dna-double-helix.html>

<https://studylib.net/doc/5786293/snp-presentation>

@@

>>>>

title: |
  | Genetic endowments
  | and wealth inequality

@@

  *Journal of Political Economy*, 2020, vol. 128, No.4

@@

    theme: [sky]

@@

    includes:
      in_header: ../sky_add_FilePointer.html
      after-body: ../toc-add.html

@@

include-before-body:
#### latex math shorthand
  file: ../MathShorthand.html 

@@

@BarthPapageorgeThom2020

@@

## SNP, polygenic score

@@

::::{.fragment}
Genome

![Chromosone...DNA](DNAMacrostructure.jpg)

::: {style="font-size: 70%;"}
[https://courses.lumenlearning.com/...](https://courses.lumenlearning.com/suny-ulster-ap1/chapter/the-nucleus-and-dna-replication)
:::
::::

@@

::: {style="font-size: 70%;"}
_[https://www.haibunda.com/...](https://www.haibunda.com/bundapedia/20230214003129-210-297249/deoxyribonucleic-acid--dna-)_
:::
::::

@@

::::{.fragment}
![Double helix and rung](DoubleHelixRung.jpg)

::: {style="font-size: 70%;"}
[https://mammothmemory.net/...](https://mammothmemory.net/biology/dna-genetics-and-inheritance/dna-base-pairing/dna-double-helix.html)
:::
::::

::::{.fragment}
![SNP](SNP.png)

::: {style="font-size: 70%;"}
_[https://studylib.net/...](https://studylib.net/doc/5786293/snp-presentation)_
:::
::::

@@

<!-- tree.nathanfriend.io -->

```{md}
.
└── Chromosome
    └── DNA
        └── Gene
            └── Genome
```

:::::{.columns}

::::{.column width="55%"}
![](genome.jpg)

::: {style="font-size: 70%;"}
[_Source: L. C. Douglas_](https://historiesofthingstocome.blogspot.com/2010/08/age-of-genome-4.html){height=5cm}
:::
::::

::::{.column width="45%"}
* Chromosomes: 1/2 each from both parents 
* DNA: Rolled up and shapes like an X
* Gene: DNA segments
* Genome: 3 billion nucletiodes (letters)
   * Letter sequence = genome sequence
   * 1.5 billion base pairs
::::

* Base pairs differ between people in less than 1% (15 million) locations  
* Single-nucleotide polymorphism (SNP) = such location
   * 2 chromosomes per person: AT-AT, GC-GC, AT-GC or 0, 2, 1 (reference = GC)

Twin studies

* Estimate how much genetic factors *collectively* matter for **explaining variations** of a trait  
   * Collectively: Unit = group level aggregation
   * Explaining variations: Variance decomposition 
* Do not reveal which SNP is correlated

. . .

Genome wide association studies (GWASs)

* Collect $J$ observable (?) SNPs of individual $i$.

. . .

Regress outcome $Y_{i}$ on each SNP using $J$ estimating equations:

$$
Y_{i}=\bfmu'\bfx_{i}+\beta_{j}SNP_{ij}+\epsilon_{ij}, \quad j=1,\dots,J.
$$

Polygenic score of $i$ for the outcome $Y$ = *"Educational attainment (EA) score"*

$$
PGS_{i}=\sum_{j=1}^{J}\tilde{\beta} SNP_{ij}
$$

* Use Bayesian LDpred procedure to correct for correlations in $\tilde{\beta}_{j}$

* Use all SNPs: Better out-of-sample results than using only SNPs with genome-wide significance $p$ value $< 5*10^{-8} =$ .0000005%

* PGS is considered to be a predictor of individual fixed effects
   * HRS sample: covariance with EA score / total schooling years variance = 10.6%
   * Edu attainment SNPs $\propto$ biological process of brain development, cognition [@Okbay2016; @Lee2018]
   * Edu attainment SNPs $\propto$ cognition SNPs [@Okbay2016]

## Interpretation

* Correlation, not causality: Outcomes &lArr; gene + environment (endogenous)
* Attenuation due to measurement errors in $\hat{\beta}_{j}$
* Linearity assumption &rarr; underestimation of gene contributions (relative to twin studies)
* Population stratification: Ethnic-group fixed effects &lArr; gene, environment  
   * Including principal components of SNP data as covariates is shown to controll geographic variations
* External validity is up to HRS ancestral composition

## Data

Health and retirement survey: Age > 50 + partners, genetic samples 2006, 2008

* All households with at least one individual with genetic European 
* "Retired households" in 1996, 1998, 2002, 2004, 2006, 2008, 2010
* Households with 1-2 members, drop same sex HHs
* Household-year observations of both members 65-75 years old

. . .

2590 HHs, 5701 HH-year observations (Table 1)
